Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-01-31 paper/sec/formalization.tex
Paper   - Lots of edits   - Add formalized version of algorithm

\section{Formalization}
\todo{Overview of the two AFP entries. Include main statement of correctness for the CVM algorithm.}

>>>>

\begin{algorithm}[h]
\caption{Formalized Version of \cref{alg:cvm_new}}\label{alg:cvm_formalized}
\begin{isabelle_cm}
\isacommand{context}\isamarkupfalse%
\isanewline
\ \ \isakeyword{fixes}\ f\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ real\ \isakeyword{and}\ n\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ nat\isanewline
\ \ \isakeyword{assumes}\ f{\isacharunderscore}{\kern0pt}range{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}f\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{1}}{\isacharslash}{\kern0pt}{\isadigit{2}}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}\ \isakeyword{and}\ n{\isacharunderscore}{\kern0pt}gt{\isacharunderscore}{\kern0pt}{\isadigit{0}}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}n\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
\isakeyword{begin}\isanewline
\isanewline
\isacommand{definition}\isamarkupfalse%
\ {\isacartoucheopen}initial{\isacharunderscore}{\kern0pt}state\ {\isacharequal}{\kern0pt}\ State\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}\ {\isadigit{1}}{\isacartoucheclose}\ %
\hfill\isamarkupcmt{Setup initial state $\chi=\emptyset$ and $p=1$.\;%
}\isanewline
\isacommand{fun}\isamarkupfalse%
\ subsample\ \isakeyword{where}\ %
\hfill\isamarkupcmt{Subsampling operation: Sample random $n f$ subset.\;%
}\isanewline
\ \ {\isacartoucheopen}subsample\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ pmf{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}set\ {\isacharbraceleft}{\kern0pt}S{\isachardot}{\kern0pt}\ S\ {\isasymsubseteq}\ {\isasymchi}\ {\isasymand}\ card\ S\ {\isacharequal}{\kern0pt}\ n\ {\isacharasterisk}{\kern0pt}\ f{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
\isanewline
\isacommand{fun}\isamarkupfalse%
\ step\ \isakeyword{where}\ %
\hfill\isamarkupcmt{Loop body.\;%
}\isanewline
\ \ {\isacartoucheopen}step\ a\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
\ \ \ \ b\ {\isasymleftarrow}\ bernoulli{\isacharunderscore}{\kern0pt}pmf\ p{\isacharsemicolon}{\kern0pt}\isanewline
\ \ \ \ let\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ {\isacharparenleft}{\kern0pt}if\ b\ then\ {\isasymchi}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}a{\isacharbraceright}{\kern0pt}\ else\ {\isasymchi}\ {\isacharminus}{\kern0pt}\ {\isacharbraceleft}{\kern0pt}a{\isacharbraceright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
\isanewline
\ \ \ \ if\ card\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ n\ then\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
\ \ \ \ \ \ {\isasymchi}\ {\isasymleftarrow}\ subsample\ {\isasymchi}{\isacharsemicolon}{\kern0pt}\isanewline
\ \ \ \ \ \ return{\isacharunderscore}{\kern0pt}pmf\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ {\isacharparenleft}{\kern0pt}p\ {\isacharasterisk}{\kern0pt}\ f{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isanewline
\ \ \ \ {\isacharbraceright}{\kern0pt}\ else\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
\ \ \ \ \ \ return{\isacharunderscore}{\kern0pt}pmf\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\isanewline
\ \ \ \ {\isacharbraceright}{\kern0pt}\isanewline
\ \ \ {\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
\isanewline
\isacommand{fun}\isamarkupfalse%
\ run{\isacharunderscore}{\kern0pt}steps\ \isakeyword{where}\ %
\hfill\isamarkupcmt{Iterate loop over stream \isa{xs}.\;%
}\isanewline
\ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}steps\ xs\ {\isacharequal}{\kern0pt}\ foldM{\isacharunderscore}{\kern0pt}pmf\ step\ xs\ initial{\isacharunderscore}{\kern0pt}state{\isacartoucheclose}\isanewline
\isacommand{fun}\isamarkupfalse%
\ estimate\ \isakeyword{where}\ 
{\isacartoucheopen}estimate\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ card\ {\isasymchi}\ {\isacharslash}{\kern0pt}\ p{\isacartoucheclose}\isanewline
\isacommand{fun}\isamarkupfalse%
\ run{\isacharunderscore}{\kern0pt}algo\ \isakeyword{where}\ %
\hfill\isamarkupcmt{Run algorithm and estimate.\;%
}\isanewline
\ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}algo\ xs\ {\isacharequal}{\kern0pt}\ map{\isacharunderscore}{\kern0pt}pmf\ estimate\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}steps\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
\isanewline
\isacommand{theorem}\isamarkupfalse%
\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
\isanewline
\isacommand{theorem}\isamarkupfalse%
\ correctness{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}\isanewline
\isacommand{end}
\end{isabelle_cm}
\end{algorithm}
\section{Formalization}
This entry is accompanied by two \todo{sessions?}.

The first is a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
The formalization of the algorithm is presented in \cref{alg:cvm_formalized}.
We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
\todo{Explain: card}
\todo{Explain: map-pmf}
\todo{Explain: pmf-of-set}
\todo{Explain: return-pmf}

### 2025-01-31 paper/sec/invariants.tex
Paper   - Lots of edits   - Add formalized version of algorithm

\section{Probabilistic Invariants}\label{sec:invariants}
In this section, we will derive our new technique using \cref{alg:cvm} as an example. 
For this it is best to consider \cref{alg:cvm} without line 11, i.e., we are going to ignore the second check whether the subsampling step succeeded, instead the algorithm continues even though the buffer may still contain $n$ elements.
The total variational distance between these two variants of the algorithms is at most $\frac{\delta}{2}$.
This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction of $\frac{\delta}{2}$.
Considering the modified version, makes the discussion easier, because we do not have to worry about $\bot$.

@@

Let us consider verifying that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \int_\Omega \int_{\Ber(p)} p^{-1} I\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
\]
Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, but it is useful to remember that these depend on $\sigma$.
To see, that the right-hand-side is equal to $1$, it is usefule to consider the cases, where $a=s$ and the converse seperately.
In the first case the right-hand-side is equal to $1$, can be seen directly, note we assume $p \in (0;1]$.
In the second case it follows from the induction hypothesis.
(In particular, the term in the inner integral is constant with respect to $\tau$.)

The same is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
Then
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
    \bigint_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{I(s \in \tau)}{p/2} \, d \tau\right)}{\frac{I(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
\]
Note that the true- and false-case both evaluate to the same value: $p^{-1} I(s \in \chi)$.
If $s \notin \chi$ both sides are $0$, because the subsampling operation returns a subset of $\chi$.
If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} I(s \in \chi)$.
Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.

@@

Normally, we would use Chernoff bounds for indepenent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
But they are not useful in our case, because the random variables in the case of the CVM algorithm, for example $p^{-1} X_s$, are not independent.
An alternative is the Cram\'{e}r-Chernoff method, which is a very general method to obtain tail bounds for any random variable.
It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t)$ is the moment generating function of $X$: $M(t) := \expect [\exp(t X)]$.
Note that, sometimes the moment generating function is only defined for some values of $t$ (when the corresponding integral exists).
This is typically an interval including $0$.
In which case the inequality is only applicable for those values.
Because the probability space, representing the state of the CVM algorithm is finite, in our case $M(t)$ will always be defined.
Of course, it is also possible to obtain lower tail bounds $P(X \leq a)$ which just requires estimates for the $M(t)$ for $t < 0$, instead of $t > 0$.

This means for our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
\[
  \expect [\exp( t p^{-1} |\chi| )] = \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right]
\]
for $h(x) = \exp(tx)$.
At this point, it was of course tempting to see, whether the proof for \cref{eq:indicator_eq} can also be extended to establish bounds for the above.
When we tried, we got the following result:
\[
  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|}
\]
for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
Note that the exponetial function is convex.
So this does not work directly.
However, when we instead try to derive tail bounds for the random variable $I(p \geq q) p^{-1} |\chi|$ instead. 
We end up with the condition that $h$ needs to be no-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$.
This then leads to tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
Of course this requires, separately estimating $\prob(p < q)$, but this is fairly standard trick.
In particular the original proof by Chakraborty et al.~\cite{chakraborty2022} use the same trick.
They use $q = \frac{n}{4 |A|}$ and we could use the same cut-off.

Note that the formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof in its appendix.
The verified algorithm is actually a generalization of the above, where the subsampling probability can be some $f \in [\frac{1}{2};e^{-1/12}]$.
The algorithm presented by Chakraborty et al.~\cite{chakraborty2022}, replicated in the introduction (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are very elementary.
The main point we wanted to make in this section is that, it is possible to derive useful and general probabilistic invariants, by considering expectations of functions of the state, using recursion or induction over the algorithm itself.
As far as we know this method of establishing tail bounds for randomized algorithms is new.

A more interesting consequence of our new method is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
We think that this new variant would not be possible to verify, without the new approach presented here.

>>>>

\section{Probabilistic Invariants}\label{sec:invariants}
In this section, we will derive our new technique using \cref{alg:cvm} as an example.

Let us briefly, review the algorithm. The algorithm's state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$.
The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely -- if the current stream element is already in the buffer -- removing them with probability $(1-p)$.
If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returing $\bot$.
After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.

For the discussion in this section, it is best to consider \cref{alg:cvm} without line 11, i.e., we are going to ignore the second check whether the subsampling step succeeded, instead the algorithm continues even though the buffer may still contain $n$ elements.
The total variational distance between these two variants of the algorithms is at most $\frac{\delta}{2}$.
This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction of $\frac{\delta}{2}$.
Considering the modified version, makes the discussion easier, because we do not have to worry about $\bot$.

@@

Let us consider verifying that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \int_\Omega \int_{\Ber(p)} p^{-1} I\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
\]
Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these depend on $\sigma$.
To see, that the right-hand-side is equal to $1$, it is useful to consider the cases, where $a=s$ and the converse seperately.
In the first case the right-hand-side is equal to $1$, can be seen directly, note we assume $p \in (0;1]$.
In the second case it follows from the induction hypothesis.
(In particular, the term in the inner integral is constant with respect to $\tau$.)

The same is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
Then
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
    \int_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{I(s \in \tau)}{p/2} \, d \tau\right)}{\frac{I(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
\]
Note that the true- and false-case both evaluate to the same value: $p^{-1} I(s \in \chi)$.
If $s \notin \chi$ both sides are $0$, because the subsampling operation returns a subset of $\chi$.
If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} I(s \in \chi)$.
Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.

@@

Normally, we would use Chernoff bounds for indepenent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
But they are not useful in our case, because the random variables in the case of the CVM algorithm, for example $p^{-1} X_s$, are not independent.
An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t)$ is the moment generating function of $X$: $M(t) := \expect [\exp(t X)]$.
Note that, sometimes the moment generating function is only defined for some values of $t$ (when the corresponding integral exists).
This is typically an interval including $0$.
In which case the inequality is only applicable for those values.
Because the probability space, representing the state of the CVM algorithm is finite, in our case $M(t)$ will always be defined.
Of course, it is also possible to obtain lower tail bounds $P(X \leq a)$ which just requires estimates for the $M(t)$ for $t < 0$, instead of $t > 0$.

This means for our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
\[
  \expect [\exp( t p^{-1} |\chi| )] = \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right]
\]
for $h(x) = \exp(tx)$.
At this point, it was of course tempting to see, whether the proof for \cref{eq:indicator_eq} can also be extended to establish bounds for the above.
When we tried, we got the following result:
\[
  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|}
\]
for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
Note that the exponetial function is convex.
So this does not work directly.
However, when we instead try to derive tail bounds for the random variable $I(p \geq q) p^{-1} |\chi|$.
We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which successfully leads to tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
Of course, we also have to separately estimate $\prob(p < q)$.
We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}.
They use $q = \frac{n}{4 |A|}$ and we could use the same cut-off.

Note that the formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof in its appendix.
The verified algorithm is actually a generalization of the above, where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$.
The algorithm presented by Chakraborty et al.~\cite{chakraborty2022}, replicated in the introduction (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are very elementary.
The main point we wanted to make in this section is that, it is possible to derive useful and general probabilistic invariants, by considering expectations of functions of the state, using recursion or induction over the algorithm itself.
As far as we know this method of establishing tail bounds for randomized algorithms is new.

When we look at the subsampling step, our proof imposes the following condition on the subsampling operation:
\begin{equation}\label{eq:subsample_condition}
  \int_{\mathrm{subsample}(\chi)} \left(\prod_{s \in S} g(I(s \in \tau)) \, d \tau\right) \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
\end{equation}
for all non-negative functions $g$.
An interesting consequence of that, is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
We think that this new variant would not be possible to verify, without the new approach presented here.

### 2025-01-31 paper/sec/negdep.tex
Paper   - Lots of edits   - Add formalized version of algorithm

The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
Why we are interested in the lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
The latter is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
Intuitively, this is interesting because it opens up an avenue to modifying the subsampling step in \cref{alg:cvm}.
Instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, which has the disadvantage that there is a small possibility that none of the elements gets discarded, we could instead choose a random $nf$-subset, where $\frac{1}{2} \leq f < 1$, such that $nf$ is integer, in the subsampling step.
For such a subsampling step, we can derive using \cref{le:neg_assoc_prod}:
\begin{equation}\label{eq:subsample_with_n_subsets}
  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(I(s \in \chi)) \leq \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(I(s \in \chi)) = \prod_{s \in S}  \expect_{\Ber(f)} g \textrm{.}
\end{equation}
for any non-negative $g$ and $S \subseteq \chi$. Note that $g$ is only evaluated at two points and is thus either non-increasing or non-decreasing.Also, if $S$ is a singleton, the inequality becomes an equality.

To understand, why the above is useful.
When we want to establish the moment generating function, for the random variable $p^{-1} |\chi|$ --- i.e. the result --- inductively, we need to approximate the then-part of the following term, with the else-part.
\[
  \ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \prod_{s \in S} h((pf)^{-1}I(s \in \tau)) \, d \tau\right)}{\prod_{s \in S} h(p^{-1}I(s \in \tau))}
\]
and the inequality (\cref{eq:subsample_with_n_subsets}) is exactly, what we need.
This means the proof for the modified algorithm in \cref{sec:invariants} works just as well with the new subsampling method.

The resulting algorithm is depicted in \cref{alg:cvm_new}.
Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
And based on our discussion from the previous section it is unbiased.

\begin{algorithm}[h!]
	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$}
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment Select a random $nf$ subset of $\chi$
      \State $p \gets pf$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}

>>>>

The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
Why we are interested in the lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
The latter is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
Intuitively, this is interesting because it opens up an avenue to modifying the subsampling step in \cref{alg:cvm}.
Instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, which has the disadvantage that there is a small possibility that none of the elements gets discarded, we could instead choose a random $nf$-subset, where $\frac{1}{2} \leq f < 1$, such that $nf$ is integer, in the subsampling step.
For such a subsampling step, we can derive using \cref{le:neg_assoc_prod}:
\begin{equation}\label{eq:subsample_with_n_subsets}
  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(I(s \in \chi)) \leq \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(I(s \in \chi)) = \prod_{s \in S}  \expect_{\Ber(f)} g \textrm{.}
\end{equation}
for any non-negative $g$ and $S \subseteq \chi$. Note that $g$ is only evaluated at two points and is thus either non-increasing or non-decreasing.
Also, if $S$ is a singleton, the inequality becomes an equality.

Let us recall that this is the exact condition we needed for the subampling step (\cref{eq:subsample_condition}).
This means the proof for the modified algorithm in \cref{sec:invariants} works just as well with the new subsampling method.

The resulting algorithm is depicted in \cref{alg:cvm_new}.
\begin{algorithm}[t]
	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$}
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment Select a random $nf$ subset of $\chi$
      \State $p \gets pf$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}%
Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
And based on our discussion from the previous section it is also an unbiased algorithm, i.e., the expectation of the result is exactly $|A|$.
